Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2018-12-04 Manuscript/manuscript.Rmd
manuscript edits

Most accepted methods for predicting TAD boundaries consider performing a classification algorithm by binning the genome into equal-sized bins and labeling them as either overlapping with a flanked TAD boundary or not. However, a pervasive issue becomes the introduction of highly imbalanced classes. Due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. Therefore, the minority class of genomic bins with TAD boundaries in them becomes proportionally smaller than that of the majority class of genomic bins with no TAD boundaries. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple].  However, a clear method for creating balanced classes remains evasive and largely dependent on the problem at hand. For that reason we evaluated 3 widely accepted class balancing techniques, in addition to no balancing. They included random under-sampling, random over-sampling, and synthetic minority over-sampling technique (SMOTE).

Another problem lies in identifying the feature space of subsequent classification algorithms.  A variety of positional characteristics of the genomic features can be used to describes the relationship between TAD boundaries and genomic elements, but it is unclear which is best. The use of distances (in base pairs) from genomic bins to regions of functional genomic elements has largely been ignored. However, these distance characteristics offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a genomic bin containing a TAD boundary--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance on imbalanced data. More robust metrics like the F1-score and Matthew's Correlation Coefficient are more suitable for this type of analysis and can highlight problems associated with applying predictive models on imbalanced data.

@@

Topologically associating domain data for the GM12878 cell line was obtained using the arrowhead algorithm from the Juicer toolbox (https://github.com/aidenlab/juicer). Publically available domain data was obtained from in situ Hi-C contact matrices at 5kb resolution [@rao20143d]. Additional domain data was obtained from Dali et al. at 25kb, 50kb, and 100kb resolutions [@dali2017critical].

Identified TAD boundaries were represented by their location in the genome (hg19 human genome assembly), including chromosome, start and end coordinates for chromosomes 1 through 22. The start and end coordinates were concatenated, sorted, and unique coordinates representing the borders demarcating TADs were obtained.

To establish the data for modeling, the genome was binned into equally sized intervals (Figure 1). This was performed at the same resolution of the contact matrix that the domain data was obtained from. For example, for the domain data obtained from a 5kb resolution contact matrix, the genome was subsequently binned into 5kb intervals. The same processes was repeated for each of the other three resolutions. Next, the boundary coordinates were flanked on either side by a unit corresponding to the resolution of the particular domain data. Again, siting the domain data obtained from a 5kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 5kb for a total width around each boundary point of 10kb. The same process was repeated for the other three resolutions. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

Annotation data, in the form of functional genomic elements from ChIP-seq experiments were obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium. The annotations consisted of a variety of different chromatin states, histone marks, and DNase I hypersensitive sites. Table 1 presents the total list of genomic  annotations considered in this study and the sources of downloadable files. Each annotation was represented by their location throughout the genome via chromosomal coordinates. An example of a specific annotation, which is labeled as a functional genomic element, is presented in Figure 2. It is represented by the green segments that overlap with the genomic bins.

##Feature Engineering

Using each genomic element, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries in downstream modelling. Figure 2 presents an illustration of each type of predictor that was considered. The predictors consisted of two different types of overlaps, and a distance type. They are described in detail below:

@@

[Mention transformations used]

>>>>

Most accepted methods for predicting TAD boundaries consider performing a classification algorithm by binning the genome into equal-sized bins and labeling them as either overlapping with a flanked TAD boundary or not. However, a pervasive issue becomes the introduction of highly imbalanced classes. Due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. Therefore, the minority class of genomic bins with TAD boundaries in them becomes proportionally smaller than that of the majority class of genomic bins with no TAD boundaries. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple].  A clear method for creating balanced classes remains evasive and largely dependent on the problem at hand.

Another problem lies in identifying the feature space of subsequent classification algorithms.  A variety of positional characteristics of the genomic features can be used to describe the relationship between TAD boundaries and genomic elements, but it is unclear which is best. The use of distances (in base pairs) from genomic bins to regions of functional genomic elements has largely been ignored. However, these distance characteristics offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a genomic bin overlapping with a TAD boundary region--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance on imbalanced data. More robust metrics like the F1-score and Matthew's Correlation Coefficient are more suitable for this type of analysis and can highlight problems associated with applying predictive models on imbalanced data.

@@

Replicate Hi-C data for the GM12878 cell line were downloaded from GEO GSE63525 (Experiments HIC001 to HIC029) [@rao20143d]. The chromosomal coordinates of topologically associating domains throughout the genome were obtained using the arrowhead algorithm from the Juicer toolbox provided by the Aiden Laboratory (https://github.com/aidenlab/juicer) [@durand2016juicebox]. Using the juicer toolbox, we were able to call TADs on 4 different contact matrix resolutions, which included 5 kb, 25 kb, 50 kb, and 100 kb. Identified TAD boundaries were represented by their location on the linear genome (hg19 human genome assembly), including start and end coordinates for chromosomes 1 through 22 (Figure 18 A). The start and end coordinates of all hierarchical TADs were concatenated, sorted, and unique boundary points representing the borders demarcating TADs were obtained (Figure 18 B).

##MODEL CONSTRUCTION

To establish the data for modeling, the genome was binned into equally sized intervals (Figure 18 C). This was performed at the same resolution of the contact matrix that the domain data was obtained from. For example, for the domain data obtained from a 5kb resolution contact matrix, the genome was subsequently binned into 5kb intervals. The same processes was repeated for each of the other three resolutions. Next, the boundary coordinates were flanked on either side by a unit corresponding to the resolution of the particular domain data. For example, for the domain data obtained from a 5kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 5kb for a total width around each boundary point of 10kb. The same process was repeated for the other three resolutions. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

Annotation data, in the form of functional genomic elements from ChIP-seq experiments were obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium (Table 1). The annotations consisted of a variety of different chromatin states, histone marks, and DNase I hypersensitive sites. Each annotation was represented by their location throughout the genome via chromosomal coordinates.

###FEATURE ENGINEERING USING GENOMIC ANNOTATIONS

Using each genomic element, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries in downstream modelling (Figure 2). The predictors consisted of two different types of overlaps, and a distance type. They are described in detail below:

@@
